---
title: "Activity 4: Data visualization 2: Customizing plots"
subtitle: Jan. 14th, 2026, Calvin Munson
format: markdown_github
editor: visual
---

# Overview

Welcome! Today's activity is all about the basics of customizing plots. We will continue using `ggplot2.`

You will submit two outputs for this activity:

1.  A **PDF** of a rendered Quarto document with all of your R code (see below). Please include all of the code that appears in this document, in addition to adding your own code in the "Q#" sections.

2.  A plot and research question from the final section

A note about submitting code: 

As a general practice, it is good to **use headers to denote which section of the homework** you are in and **add in your own comments about what each line is actually doing** - this is excellent practice for organizing your own big R scripts in the future. The biggest way that you will code for probably the first few years of your coding career will consist of you revisiting old R scripts and copying/pasting code and/or adapting that code to your current problems. I promise you, your future self will be thrilled with your current self if you take the extra few minutes to 1) organize your code in a nice way and 2) provide detailed comments about what each line is doing (as well as what questions you have about it). 

Also, don’t be afraid to use lots of blank lines! It is MUCH nicer for your future self to look at code that has a few blank lines in between sections of code than a bajillion lines in a row of straight code.

From here on, please put headers (either as comments in the code or as Quarto headers) to denote which question you are answering (e.g. "Q1:") and precede your code with comments that describe what that code does.

# 1) ____ Data

## Setting up

First, let's load our packages. You should already have the `here` and `ggplot2` packages installed; read them in from the library in your code.

```{r setup, warning = FALSE, message = FALSE}
# install.packages("ggplot2")
library(ggplot2)
library(here)



```


# 2) CO2 Uptake in Grass Plants Data

Next, it's time to practice what you've learned again! R has a number of built-in datasets. The dataset `CO2` provides the data from an experiment on the cold tolerance of a grass species. Familiarize yourself with the dataset by reading the help page for this dataset

```{r}
# Putting a "?" in front of a function (or built-in dataset) will automatically bring up its help page!
?CO2
```

## Q2.1: Write a research question that you would like to answer

Once you have a sense of what is in the dataset, write a research question that you would like to answer. There are many options with varying complexity!

## Q2.2: Choose a graph to answer your question

What graph would be appropriate to answer this question? Think about the kinds of graphs we've used so far and whether your variables are continuous or categorical.

## Q2.3: Create your graph



